---
title: "Building Aggregation Maps From Row and Column Labels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building Aggregation Maps From Row and Column Labels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction

Aggregation maps are helpful when aggregating products and industries
in rows and columns of PSUT matrices (**R**, **U**, **V**, and **Y**).
This vignette demonstrates ways to use row and column names 
from PSUT matrices to create an aggregation map.

To provide an example,
this vignette uses data from one of the CL-PFU Database products, 
namely the producte named "psut" from v0.9 of the database, 
which is a `matsindf` data frame containing PSUT matrices for the USA, 1960--2019.


## Load PSUT data

To work with the CL-PFU database as shown in this vignette, 
be sure you have access to several packages.

```{r setup, message=FALSE}
library(dplyr)
library(magrittr)
library(matsbyname)
library(pins)
library(Recca)
library(readxl)
```

To load the data, use code like this:

```{r}
pinboard <- pins::board_folder("~/Dropbox/Fellowship 1960-2015 PFU database/OutputData/PipelineReleases")
df <- pinboard |>
  # Load USA data from v0.9 of the database
  pins::pin_read(name = "psut_usa", version = "20230220T223535Z-35e3e")
dplyr::glimpse(df)
```


## Extract row and column names

`Recca::get_all_products_and_industries()`
will find the product and industry names for each row of the matsindf data frame.
The results are placed in two new columns of the data frame
named (by default) "Product.names" and "Industry.names".

```{r}
df_with_prods_inds <- df |> 
  Recca::get_all_products_and_industries()
dplyr::glimpse(df_with_prods_inds)
```

Product and industry names can be quite different
from row to row in the data frame, 
because there are many new energy carriers (products)
when the Last.stage is "Useful" compared to "Final".

```{r}
# Check a row with "Final" energy as the Last.stage.
df_with_prods_inds |> 
  dplyr::filter(Energy.type == "E", 
                Last.stage == "Final", 
                Year == 1960, 
                IEAMW == "IEA") |> 
  magrittr::extract2("Product.names")

# Check a row with "Useful" energy as the Last.stage.
df_with_prods_inds |> 
  dplyr::filter(Energy.type == "E", 
                Last.stage == "Useful", 
                Year == 1960, 
                IEAMW == "IEA") |> 
  magrittr::extract2("Product.names")
```

Similarly, industry names may be quite different when Last.stage is “Final” or “Useful”.

```{r}
# Check a row with "Final" energy as the Last.stage.
df_with_prods_inds |> 
  dplyr::filter(Energy.type == "E", 
                Last.stage == "Final", 
                Year == 1960, 
                IEAMW == "IEA") |> 
  magrittr::extract2("Industry.names") |> 
  unlist()

# Check a row with "Useful" energy as the Last.stage.
df_with_prods_inds |> 
  dplyr::filter(Energy.type == "E", 
                Last.stage == "Useful", 
                Year == 1960, 
                IEAMW == "IEA") |> 
  magrittr::extract2("Industry.names") |> 
  unlist()
```


## Get lists of unique product and industry names

To obtain lists of unique product and industry names, 
one can use `unique()`.

```{r}
# Unique products in the data frame
df_with_prods_inds$Product.names |> 
  unlist() |> 
  unique()

# Unique industries in the data frame
df_with_prods_inds$Industry.names |> 
  unlist() |> 
  unique()
```


## Aggregate by parts of row and column labels

`Recca::get_all_products_and_industries()` 
can also select only parts of the row and column label. 
Knowing that we (eventually) want to aggregate by the nouns, 
we can extract only the nouns (prefixes) from the row and column labels,
as shown by the code below.
Extracting by a piece of the row and column names is
significantly slower than extracting the entire row or column names.

```{r}
df_with_prods_inds_nouns <- df |>
  # Restrict years to reduce execution time.
  dplyr::filter(Year == 1960) |>
  Recca::get_all_products_and_industries(
    piece = "noun",
    inf_notation = TRUE,
    notation = list(RCLabels::bracket_notation, RCLabels::arrow_notation))
dplyr::glimpse(df_with_prods_inds_nouns)
```

Again, one can use `unique()` to find the nouns required for aggregation maps.
These lists are shorter than 
lists of whole names of products and industries, because 
there are fewer unique nouns than unique row and column labels.

```{r}
# Unique products (nouns only) in the data frame
df_with_prods_inds_nouns$Product.names |> 
  unlist() |> 
  unique()

# Unique industries (nouns only) in the data frame
df_with_prods_inds_nouns$Industry.names |> 
  unlist() |> 
  unique()
```

The lists of nouns can be used to create aggregation maps, as shown in the next section.


## Construct an aggregation map

An aggregation map is 
a description of the way aggregation is to be performed.
In `R` terms, 
an aggregation map is simply a named list 
where values are the row or column names (or pieces of names) to be aggregated and
names are the new names for the aggregated rows or columns. 

For example,

```{r}
prod_agg_map_example <- list(
  `Coal & coal products` = c("Hard coal (if no detail)", 
                             "Brown coal (if no detail)",
                             "Anthracite",
                             "Coking coal",
                             "Other bituminous coal",
                             "Sub-bituminous coal"), 
  `Oil & oil products` = c("Crude/NGL/feedstocks (if no detail)", 
                           "Crude oil", 
                           "Natural gas liquids", 
                           "Refinery feedstocks", 
                           "Additives/blending components", 
                           "Other hydrocarbons")
)
```

Aggregation maps can also be constructed from a data frame with “Many” and “Few” columns.

```{r}
prod_agg_table_example <- tibble::tribble(
  ~Many, ~Few, 
  "Hard coal (if no detail)", "Coal & coal products",
  "Brown coal (if no detail)", "Coal & coal products",
  "Anthracite", "Coal & coal products",
  "Coking coal", "Coal & coal products",
  "Other bituminous coal", "Coal & coal products",
  "Sub-bituminous coal", "Coal & coal products",
  "Crude/NGL/feedstocks (if no detail)", "Oil & oil products",
  "Crude oil", "Oil & oil products",
  "Natural gas liquids", "Oil & oil products",
  "Refinery feedstocks", "Oil & oil products",
  "Additives/blending components", "Oil & oil products",
  "Other hydrocarbons", "Oil & oil products"
)
prod_agg_table_example

matsbyname::agg_table_to_agg_map(prod_agg_table_example,
                                 many_colname = "Many", 
                                 few_colname = "Few")
```

Aggregation tables can be created in an Excel file and
read into `R` with the `readxl` package. 
Example aggregation tables can be found in the 
"Aggregation_maps.xlsx" file in the `PFUDatabase` package.

```{r}
prod_agg_map <- system.file("extdata", "Aggregation_Tables.xlsx",  
                            package="PFUDatabase") |> 
  readxl::read_excel(sheet = "product_aggregation") |> 
  matsbyname::agg_table_to_agg_map(many_colname = "Many", 
                                   few_colname = "Few")
prod_agg_map

ind_agg_map <- system.file("extdata", "Aggregation_Tables.xlsx", 
                              package="PFUDatabase") |> 
  readxl::read_excel(sheet = "industry_aggregation") |> 
  matsbyname::agg_table_to_agg_map(many_colname = "Many", 
                                   few_colname = "Few")
ind_agg_map
```


## Perform the aggregations

There are two `matsbyname` functions that assist with aggregations: 
`matsbyname::aggregate_byname()` and
`matsbyname::aggregate_pieces_byname()`.
A third function (`matsbyname::aggregate_to_pref_suff_byname()`) 
has been deprecated in favor of `matsbyname::aggregate_pieces_byname()`.

Knowing that the `matsbyname` package can aggregate
by any piece of a row or column label
(including prefix, suffix, noun, or object of a preposition),
aggregation maps can be constructed 
by focusing only on the nouns (prefixes) 
of the row and column names.

To aggregate the matrices in the `df` object, 
one can use code like the following example.

```{r}
aggregated <- df |> 
  # Restrict to a single year to reduce execution time.
  dplyr::filter(Year == 1960) |> 
  dplyr::mutate(
    # Aggregate R
    R_agg = .data[["R"]] |>
      matsbyname::aggregate_pieces_byname(
        piece = "noun",
        margin = "Industry",
        inf_notation = TRUE,
        notation = list(list(RCLabels::bracket_notation, RCLabels::arrow_notation)),
      ) |>
      matsbyname::aggregate_pieces_byname(
        piece = "noun",
        margin = "Product",
        inf_notation = FALSE,
        notation = RCLabels::bracket_notation,
        aggregation_map = list(prod_agg_map)
      ),
    # Aggregate U_feed
    U_feed_agg = .data[["U_feed"]] |>
      matsbyname::aggregate_pieces_byname(
        piece = "noun",
        margin = "Product",
        inf_notation = FALSE,
        notation = RCLabels::bracket_notation,
        aggregation_map = list(prod_agg_map)
      ) |>
      matsbyname::aggregate_pieces_byname(
        piece = "noun",
        margin = "Industry",
        inf_notation = TRUE,
        notation = list(list(RCLabels::bracket_notation, RCLabels::arrow_notation)),
        aggregation_map = list(ind_agg_map)
      ),
    # Aggregate U_EIOU
    U_EIOU_agg = .data[["U_EIOU"]] |>
      matsbyname::aggregate_pieces_byname(
        piece = "noun",
        margin = "Product",
        inf_notation = FALSE,
        notation = RCLabels::bracket_notation,
        aggregation_map = list(prod_agg_map)
      ) |>
      matsbyname::aggregate_pieces_byname(
        piece = "noun",
        margin = "Industry",
        inf_notation = TRUE,
        notation = list(list(RCLabels::bracket_notation, RCLabels::arrow_notation)),
        aggregation_map = list(ind_agg_map)
      ),
    # Create aggregated U matrix
    U_agg = matsbyname::sum_byname(.data[["U_feed_agg"]], .data[["U_EIOU_agg"]]),
    # Create aggregated r_EIOU matrix
    r_EIOU_agg = matsbyname::quotient_byname(.data[["U_EIOU"]], .data[["U"]]) |>
      matsbyname::replaceNaN_byname(),
    # Aggregate V
    V_agg = .data[["V"]] |>
      matsbyname::aggregate_pieces_byname(
        piece = "noun",
        margin = "Industry",
        inf_notation = TRUE,
        notation = list(list(RCLabels::bracket_notation, RCLabels::arrow_notation)),
      ) |>
      matsbyname::aggregate_pieces_byname(
        piece = "noun",
        margin = "Product",
        inf_notation = FALSE,
        notation = RCLabels::bracket_notation,
        aggregation_map = list(prod_agg_map)
      ),
    # Aggregate Y
    Y_agg = .data[["Y"]] |>
      matsbyname::aggregate_pieces_byname(
        piece = "noun",
        margin = "Product",
        inf_notation = FALSE,
        notation = RCLabels::bracket_notation,
        aggregation_map = list(prod_agg_map)
      ) |>
      matsbyname::aggregate_pieces_byname(
        piece = "noun",
        margin = "Industry",
        inf_notation = TRUE,
        notation = list(list(RCLabels::bracket_notation, RCLabels::arrow_notation)),
        aggregation_map = list(ind_agg_map)
      )
  )
```

The aggregated matrices are simpler than the original matrices, as expected.

```{r}
aggregated$R_agg[[1]]

aggregated$U_feed_agg[[1]]

aggregated$U_EIOU_agg[[1]]

aggregated$U_agg[[1]]

aggregated$V_agg[[1]]

aggregated$Y_agg[[1]]
```


## Conclusion

Through judicious use of several helpful functions
(`Recca::get_all_products_and_industries()`, `matsbyname::agg_table_to_agg_map()`, and
`matsbyname::aggregate_pieces_byname()`),
one can aggregate PSUT matrices when they are arranged in a matsindf data frame.










