---
title: "PFUDatabase Demo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PFUDatabase Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dplyr)
library(PFUDatabase)
library(pins)
library(Recca)
```


## Introduction

`PFUDatabase` is a package that runs a `targets` pipeline
and stores data in Dropbox using the `pins` package.

To retrieve data, one needs

* Access to the Dropbox folder containing pipeline releases.
* A valid pin name.

One valid pin name is "psut" and a valid pin release is "20220828T174526Z-60a07".
"20220828T174526Z-60a07" contains data for all countries (155) and all years (1960--2019).

```{r}
folder <- PFUSetup::get_abs_paths()[["pipeline_releases_folder"]]
folder
# version <- "20220828T174526Z-60a07" # All countries
version <- "20220909T125050Z-e5a95" # Only GHA
psut_mats <- folder %>% 
  pins::board_folder(versioned = TRUE) %>% 
  pins::pin_read("psut", version = version)
```


## Data structure

`psut_mats` is an `R` data frame. 
Let's look at its contents.

```{r}
dplyr::glimpse(psut_mats)
```

The metadata columns contain the following values.

```{r}
psut_mats$Country %>% unique()
psut_mats$Method %>% unique()
psut_mats$Energy.type %>% unique()
psut_mats$Last.stage %>% unique()
c(psut_mats$Year %>% min(), psut_mats$Year %>% max())
psut_mats$IEAMW %>% unique()
```


## Example ECC matrices

Let's look at a few of matrices using Ghana 1971 as an example.
Simple matrices can be found in the muscle work data frames.

```{r}
GHA1971_MW <- psut_mats %>% 
  dplyr::filter(Country == "GHA", Year == 1971, Energy.type == "E", 
                Last.stage == "Final", IEAMW == "MW")
dplyr::glimpse(GHA1971_MW)
GHA1971_MW$R[[1]]
GHA1971_MW$U[[1]]
GHA1971_MW$V[[1]]
GHA1971_MW$Y[[1]]
```

These matrices can be dumped to an Excel file for easier browsing.

```{r}
gha_mw_file <- file.path("~", "GHA1971_MW.xlsx")
Recca::write_ecc_to_excel(GHA1971_MW, path = gha_mw_file, overwrite_file = TRUE)
```

Or a Sankey diagram can be created.

```{r}
GHA1971_MW %>% 
  Recca::make_sankey() %>% 
  magrittr::extract2("Sankey") %>% 
  magrittr::extract2(1)
```

The IEA data give rise to more complicated matrices.
This one is has the last stage being final energy.

```{r}
GHA1971_IEA_final <- psut_mats %>% 
  dplyr::filter(Country == "GHA", Year == 1971, Energy.type == "E", 
                Last.stage == "Final", IEAMW == "IEA")
gha_iea_file <- file.path("~", "GHA1971_IEA_final.xlsx")
Recca::write_ecc_to_excel(GHA1971_IEA_final, path = gha_iea_file, overwrite_file = TRUE)
```


```{r}
GHA1971_IEA_final %>% 
  Recca::make_sankey() %>% 
  magrittr::extract2("Sankey") %>% 
  magrittr::extract2(1)
```

When we push to useful energy, the matrices become more complex.

```{r}
GHA1971_IEA_useful <- psut_mats %>% 
  dplyr::filter(Country == "GHA", Year == 1971, Energy.type == "E", 
                Last.stage == "Useful", IEAMW == "IEA")
gha_iea_file <- file.path("~", "GHA1971_IEA_useful.xlsx")
Recca::write_ecc_to_excel(GHA1971_IEA_useful, path = gha_iea_file, overwrite_file = TRUE)
```

And the Sankey diagram is unsurprisingly busier,
but the matrices keep everything organized.

```{r}
GHA1971_IEA_useful %>% 
  Recca::make_sankey() %>% 
  magrittr::extract2("Sankey") %>% 
  magrittr::extract2(1)
```


## Example calculations

The following subsections illustrate example calculations.


### Matrix math

Here we calculate the value added matrix (**W**)
within the data frame using `dplyr::mutate()`.

```{r}
with_W <- GHA1971_MW %>% 
  dplyr::mutate(
    W = matsbyname::transpose_byname(GHA1971_MW$V) %>%
      matsbyname::difference_byname(GHA1971_MW$U)
  )
dplyr::glimpse(with_W)
with_W$W[[1]]
```


### Upstream swim

The PSUT framework allows upstream and downstream "swims."

Here is the final demand matrix for the IEA data.

```{r}
# View(GHA1971_IEA$Y[[1]])
```

Let's say we wanted to know the energy (at all stages)
required to provide residential energy in the country.
We can use an input-output ``upstream swim'' for that analysis.

First, we calculate all the input-output matrices.

```{r}
with_io_mats <- GHA1971_IEA_final %>% 
  Recca::calc_io_mats()
dplyr::glimpse(with_io_mats)
```

Next, we define a `Y_prime` matrix that contains only final demand from residences.

```{r}
with_Y_prime <- with_io_mats %>% 
  dplyr::mutate(
    Y_prime = Y %>% 
      matsbyname::select_cols_byname("Residential")
  )
with_Y_prime$Y_prime[[1]]
```

Now let's swim upstream from the `Y_prime` matrix to see what upstream energy 
is needed to supply Residential final demand.

```{r}
residential_ecc <- with_Y_prime %>% 
  Recca::new_Y()
dplyr::glimpse(residential_ecc)
```

The `_prime` matrices are what we want.
So we keep those and make a new Sankey diagram.

```{r}
residential_ecc_trimmed <- residential_ecc %>% 
  dplyr::select(Country, Method, Energy.type, Last.stage, Year, IEAMW,
                R_prime, U_prime, U_feed_prime, U_EIOU_prime, 
                r_EIOU_prime, V_prime, Y_prime, S_units) %>% 
  dplyr::rename(
    R = R_prime, 
    U = U_prime, 
    U_feed = U_feed_prime,
    U_EIOU = U_EIOU_prime,
    r_EIOU = r_EIOU_prime,
    V = V_prime, 
    Y = Y_prime
  ) %>% 
  dplyr::mutate(Sector = "Residential")
dplyr::glimpse(residential_ecc_trimmed)
```

Make a Sankey diagram of energy required to supply Residential demand only.

```{r}
residential_ecc_trimmed %>% 
  Recca::make_sankey() %>% 
  magrittr::extract2("Sankey") %>% 
  magrittr::extract2(1)
```

From the Sankey diagram, we can see all the ways in which Ghanaian residences used final energy in 1971.


### Efficiencies

With the Residential sector isolated, we can calculate the efficiency of energy use in residences.

```{r}
primary_industries <- IEATools::tpes_flows
primary_industries
finaldemand_sectors <- IEATools::fd_sectors
finaldemand_sectors
with_efficiencies <- residential_ecc_trimmed %>% 
  Recca::primary_aggregates(p_industries = primary_industries, pattern_type = "leading") %>% 
  Recca::finaldemand_aggregates(fd_sectors = finaldemand_sectors) %>%
  dplyr::mutate(
    eta_pf_net = as.numeric(EX.fd_net) / as.numeric(EX.p),
    eta_pf_gross = as.numeric(EX.fd_gross) / as.numeric(EX.p)
  )
dplyr::glimpse(with_efficiencies)
```


## Pipelines

Another package (`PFUAggDatabase`) 
does the above calculations (and more!) for
all countries; 
all years;
energy and exergy;
last stage final and useful;
IEA, MW, and both;
etc.
The calculations are parallelized across countries.
Despite parallelization, the calculations take several hours on a 10-core laptop, 
We will (eventually) generating the `PFUAggDatabase` on a supercomputer.


## Conclusion

In combination, the PSUT framework and the `R` packages
`matsbyname`, `matsindf`, and `Recca`
provide powerful analysis tools for energy conversion chains.